GMSE: an R package for generalised management strategy evaluation

### 2017-10-18 notebook/ms/ms.Rmd
Update to manuscript, including new function mentions

1. Management strategy evaluation (MSE) is a powerful tool for simulating all key aspects of natural resource management under conditions of uncertainty.
2. We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
3. GMSE models can be agent-based and spatially explicit, incorporating a high degree of realism through mechanistic modelling of links and feedbacks among stakeholders and with the ecosystem; additionally, user-defined sub-models can also be incorporated as functions into the broader GMSE framework. 
4. We show how GMSE simulates a social-ecological system using the example of a waterfowl population in an agricultural landscape that is adaptively managed; simulated waterfowl exploit agricultural land, causing conflict between conservation interests and the interest of food producers maximising their crop yield.
5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.

@@

GMSE builds off of the MSE framework, which includes four sub-models, each of which runs once in a single time step of the broader model (Figure 1). (1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty. This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management. (2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting on the whole landscape either one linear transect or one rectangular block at a time (during which resources might move). Sampling error from all of these mechanisms of observation generates a range of uncertainties depending on monitoring effort. (3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their pre-defined target abundance. Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories. After a suitable policy is found, (4) users can perform actions that affect resources or landscape cells. Users respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints. Once each user has found an adaptive strategy, user actions affect resources and landscape cells, feeding back into the resource sub-model.

@@

Simulations are run using the gmse() function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of gmse() is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation. Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function). An example below shows how simulations are set and interpreted.

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops [e.g., @Madsen2017]. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size. Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for 10 farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

Parameters in gmse() not listed are set to default values. By plotting the output with plot_gmse_results, simulation results can be interpreted visually.

@@

```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of 10 stakeholders. The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation. The upper right panel shows the same landscape broken down into 10 differently coloured regions, which correspond to areas of land owned by each of the 10 stakeholders. The middle left panel shows the actual abundance of resources (black solid line, and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals of a mark-recapture analysis), over time. The horizontal dotted red and blue lines show the resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively. The orange line shows the total percent yield of landscape cells. The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel. The lower left panel shows the cost of stakeholders performing actions over time, as set by the manager. The lower right panel shows the total number of actions attempted to be performed by all stakeholders over time (some actions might be unsuccessful if resources are unavailable on a stakeholder's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
                       land = sim$land, agents = sim$agents, paras = sim$paras,
                       ACTION = sim$action, COST = sim$cost);
```

@@

Only the estimates of population size from the observation model are available to the manager, so policy change at any time step is driven primarily by the deviation of the currently estimated population size from the manager's target and the actions of farmers in the previous time step. Hence, when the population size is estimated to be below (above) the manager's target, the manager increases (decreases) the cost of culling and decreases (increases) the cost of scaring. Because the manager does not know in advance how stakeholders will react to policy change, they assume a proportional response in total actions with respect to a change in cost (e.g., doubling the cost of culling will decrease stakeholder culling by $1/2$). Farmers responding to policy are interested only in minimising waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper). This is reflected in the bottom left versus right panels of Figure 2; when managers decrease culling costs relative to scaring, farmers respond with more total culling, and vice versa. Farmer decisions then affect waterfowl distribution and abundance, impacting future crop yield and policy.

Future development
===================================================

The GMSE package is under continued development to include additional features that will be of interest to conservation biologists, managers, and the general public. Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface. Additionally, future versions of GMSE will allow software users to take the place of stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function). Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, so GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework. The GMSE R package is therefore a versatile and collaborative tool with widespread applications for adaptive resource management.

>>>>

1. Management strategy evaluation (MSE) is a powerful tool for simulating all key aspects of natural resource management under conditions of uncertainty.
2. We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive decision-making management scenarios between stakeholders with competing objectives under complex social-ecological interactions and uncertainty.
3. GMSE models can be agent-based and spatially explicit, incorporating a high degree of realism through mechanistic modelling of links and feedbacks among stakeholders and with the ecosystem; additionally, user-defined sub-models can also be incorporated as functions into the broader GMSE framework. 
4. We show how GMSE simulates a social-ecological system using the example of an adaptively managed waterfowl population on an agricultural landscape; simulated waterfowl exploit agricultural land, causing conflict between conservation interests and the interest of food producers maximising their crop yield.
5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.

@@

GMSE builds off of the MSE framework (Figure 1), and the `gmse` function runs simulations using four default submodels. (1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty. This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management. (2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting across the whole landscape either one linear transect or one rectangular block at a time (during which resources might move). Sampling error from all observation types generates a range of uncertainties that depend on monitoring effort. (3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their pre-defined target abundance. Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories. After a suitable policy is found, (4) users perform actions that affect resources or landscape cells. Users respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints. Once each user has found an adaptive strategy, user actions affect resources and landscape cells, feeding back into the resource sub-model.

@@

Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation. Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the `plot_gmse_results` function, and a summary of results can be obtained using `gmse_summary`). An example below shows how simulations are set and interpreted.

@@

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of user land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with users [hereafter 'farmers', e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops [e.g., @Madsen2017]. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. We consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. In each time step, waterfowl are observed on subset of cells, then managers extrapolate from density per cell to estimate total population size. Managers then use these estimates to set costs of culling and scaring (non-lethal) waterfowl for five farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

Parameters in `gmse` not listed are set to default values. By plotting the output with `plot_gmse_results`, simulation results can be interpreted visually (manager and user decisions can also be interpreted using the `plot_gmse_effort` function).

@@

```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of five farmers The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation. The upper right panel shows the same landscape broken down into five differently coloured regions, which correspond to areas of land owned by each of the five farmers. The middle left panel shows the actual abundance of resources (black solid line), and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals), over time. The horizontal dotted red and blue lines show the resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively. The orange line shows the total percent yield of landscape cells. The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel. The lower left panel shows the cost of farmers performing actions over time, as set by the manager. The lower right panel shows the total number of actions attempted to be performed by all farmers over time (some actions might be unsuccessful if resources are unavailable on a farmer's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
                       land = sim$land, agents = sim$agents, paras = sim$paras,
                       ACTION = sim$action, COST = sim$cost);
```

@@

Only the estimates of population size from the observation model are available to the manager, so policy change at any time step is driven primarily by the deviation of the currently estimated population size from the manager's target and the actions of farmers in the previous time step. Hence, when the population size is estimated to be below (above) the manager's target, the manager increases (decreases) the cost of culling and decreases (increases) the cost of scaring. Because the manager does not know in advance how farmers will react to policy change, they assume a proportional response in total actions with respect to a change in cost (e.g., doubling the cost of culling will decrease stakeholder culling by $1/2$). Farmers responding to policy are interested only in minimising waterfowl's exploitation of their crops, so they will either cull or scare to remove the waterfowl from their land, depending on which option is more effective (i.e., cheaper). This is reflected in the bottom left versus right panels of Figure 2; when managers decrease culling costs relative to scaring, farmers respond with more total culling, and vice versa. Farmer decisions then affect waterfowl distribution and abundance, impacting future crop yield and policy.

Graphical User Interface (GUI)
===================================================

The function `gmse_gui` opens GMSE in a browser and allows simulations to be run for most `gmse` parameter options using the package 'shiny' [@Chang2017]. Figures from `plot_gmse_results` and `plot_gmse_effort`, and tables from `gmse_summary` are provided as GUI output.

Custom defined sub-models
===================================================

The function `gmse_apply` allows custom resource, observation, manager, or user sub-models to be integrated into the GMSE framework. Any type of sub-model (e.g., numerical, individual-based) is permitted by defining a function with appropriately specified inputs and outputs; where custom functions are not provided, `gmse_apply` runs default GMSE sub-models used in `gmse`. Any parameter options available in `gmse` or in custom functions can be passed directly to `gmse_apply`, thereby allowing for high flexibility in model specification. For example, a simple logistic growth function can be integrated as a resource sub-model to replace the default `resource` function.

```{r, echo = TRUE}
logistic_res_mod <- function(X_0, K = 2000, gr = 1){
    X_1 <- X_0 + gr * X_0 * ( 1 - X_0/K );
    return(X_1);
}
sim <- gmse_apply(res_mod = logistic_res_mod, X_0 = 1000, gr = 0.3, stakeholders = 5);
```

The `gmse_apply` function simulates a single GMSE time step, and therefore must be looped for simulations over multiple time steps.

Availability
===================================================

The GMSE package can be downloaded from CRAN (https://cran.r-project.org/package=GMSE) or GitHub (https://github.com/bradduthie/gmse). GMSE is open source under GNU Public License.
